---
title: "Lab 4"
date: "`r Sys.Date()`"
author: "Johannes Hedström, Mikael Montén"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A51 Bioinformatics \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\
  Institutionen för datavetenskap \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---

<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- creating the table of contents -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE, fig.width = 6, fig.height = 5)
```

```{r setup, echo=FALSE, messages=FALSE, warnings=FALSE}
suppressPackageStartupMessages({
    library(airway)
    library(DESeq2)
    library(ggplot2)
    library(org.Hs.eg.db)
    library(ALL)
    library(GEOquery)
})
```

# Question 1

*Run all the R code and reproduce the graphics. Go carefully through the R code and explain in your words what each step does.*
*_HINT_ Recall what a design/model matrix is from linear regression.*

```{r}
## ------------------------------------------------------------------------
library(GEOquery)
library(R.utils)

## ---- cache=TRUE---------------------------------------------------------
#x = getGEOSuppFiles("GSE20986")
#x

## ------------------------------------------------------------------------
#untar("GSE20986/GSE20986_RAW.tar", exdir = "data")

## ------------------------------------------------------------------------
#cels = list.files("data/", pattern = "[gz]")
#sapply(paste("data", cels, sep = "/"), R.utils::gzip)

## ------------------------------------------------------------------------
#phenodata = matrix(rep(list.files("data"), 1), ncol =1)
#class(phenodata)
#phenodata <- as.data.frame(phenodata)
#colnames(phenodata) <- c("FileName")
## ------------------------------------------------------------------------
library(affy)
path <- "C:/Users/wountain/Desktop/SCHOOL/HT24_Bioinformatics/Bioinformatics/lab 4/data"
fns <- list.celfiles(path, full.name = TRUE)
celfiles <- read.affybatch(celfile.path = fns)
celfiles$Targets <- c("iris", "retina", "retina", "iris", "retina", "iris", "choroid", "choroid", "choroid", "huvec", "huvec", "huvec")
boxplot(x = celfiles)
```



```{r}
library(RColorBrewer)
cols = brewer.pal(8, "Set1")
eset <- exprs(celfiles)
samples <- celfiles$Targets
colnames(eset)
colnames(eset) <- samples

boxplot(celfiles, col = cols, las = 2)
```


```{r}
distance <- dist(t(eset), method = "maximum")
clusters <- hclust(distance)
plot(clusters)
```



```{r, cache = TRUE}
## ------------------------------------------------------------------------
require(simpleaffy)
require(affyPLM)
celfiles.gcrma = gcrma(celfiles)
par(mfrow=c(1,2))
boxplot(celfiles.gcrma, col = cols, las = 2, main = "Post-Normalization");
boxplot(celfiles, col = cols, las = 2, main = "Pre-Normalization")
```


```{r}
#dev.off()

distance <- dist(t(exprs(celfiles.gcrma)), method = "maximum")
clusters <- hclust(distance)
plot(clusters)
```

```{r}
## ------------------------------------------------------------------------
library(limma)

phenodata <- matrix(c(list.celfiles(path),
                      list.celfiles(path),
                      c("iris", "retina", "retina", "iris", "retina", "iris", "choroid", "choroid", "choroid", "huvec", "huvec", "huvec")), ncol = 3)
colnames(phenodata) <- c("Name", "FileName", "Targets")
                    
samples <- as.factor(samples)
design <- model.matrix(~0+samples)
colnames(design) <- c("choroid", "huvec", "iris", "retina")
design
```


```{r}
contrast.matrix = makeContrasts(
              huvec_choroid = huvec - choroid, 
              huvec_retina = huvec - retina, 
              huvec_iris <- huvec - iris, 
              levels = design)

fit = lmFit(celfiles.gcrma, design)
huvec_fit <- contrasts.fit(fit, contrast.matrix)
huvec_ebay <- eBayes(huvec_fit)

library(hgu133plus2.db)
library(annotate)
probenames.list <- rownames(topTable(huvec_ebay, number = 100000))
getsymbols <- getSYMBOL(probenames.list, "hgu133plus2")
results <- topTable(huvec_ebay, number = 100000, coef = "huvec_choroid")
results <- cbind(results, getsymbols)

## ------------------------------------------------------------------------
summary(results)
```

```{r}
results$threshold <- "1"
a <- subset(results, adj.P.Val < 0.05 & logFC > 5)
results[rownames(a), "threshold"] <- "2"
b <- subset(results, adj.P.Val < 0.05 & logFC < -5)
results[rownames(b), "threshold"] <- "3"
table(results$threshold)
```


```{r}
## ------------------------------------------------------------------------
library(ggplot2)
volcano <- ggplot(data = results, 
                  aes(x = logFC, y = -1*log10(adj.P.Val), 
                      colour = threshold, 
                      label = getsymbols))

volcano <- volcano + 
  geom_point() + 
  scale_color_manual(values = c("black", "red", "green"), 
                     labels = c("Not Significant", "Upregulated", "Downregulated"), 
                     name = "Key/Legend")

volcano + 
  geom_text(data = subset(results, logFC > 5 & -1*log10(adj.P.Val) > 5), aes(x = logFC, y = -1*log10(adj.P.Val), colour = threshold, label = getsymbols)  )

```


















